---
title: "XCMS processing"
author: "Mar Garcia-Aloy"
output: 
  BiocStyle::html_document:
    toc: true
    number_sections: false
    toc_float: true
---

```{r startpoint, include = FALSE}
startpoint <- Sys.time()
```


# Preliminaries

## Parameters

```{r parameters}
polarity <- "NEG" # specify "POS" or "NEG"
int <- 500000
```


## Libraries

```{r libraries, message = FALSE}
library(RColorBrewer)
library(xcms)
library(SummarizedExperiment)
library(BioMark)
library(plotly)
```


## Coloring factors

```{r colors, eval=TRUE}
col_filter <- brewer.pal(4, name = "Set1")
names(col_filter) <- c("millipore",# red
                     "clarify",  # blue
                     "minisart", # green
                     "nonfilter"#,# viola
                     #"QCeq",    # orange
                     #"QC_scale",# yellow
                     #"QC_plate",# brown
                     #"STDs"     # pink
)  
plot(1, 1, xaxt = 'n', yaxt = 'n', bty = 'n', 
     xlab = '', ylab = '', xlim = 0:1, ylim = 0:1)
legend("topleft", legend = names(col_filter), bty = 'n', 
       col = col_filter, pch = 16, pt.cex = 2, cex = 2, ncol = 1)
```


# Data import

```{r data-import}
#  Get filenames
injections <- data.frame(
  filename = list.files(paste0("data/", polarity, "_fixed/"))
)
tmp <- strsplit(gsub(".mzData", "", injections$filename), "_")
tmp <- data.frame(do.call(rbind, tmp))
colnames(tmp) <- c("order", "solvent", "filter", "polarity", "repl")
injections <- cbind(injections, tmp)
rm(tmp)


# Read the data
data_raw <- readMSData(
  paste0("data/", polarity, "_fixed/" , injections$filename),
  pdata = as(AnnotatedDataFrame(injections),
             "NAnnotatedDataFrame"), 
  mode = "onDisk")
```

# BPC

```{r}
chr_raw <- chromatogram(data_raw, aggregationFun = "max")
plot(chr_raw, col = paste0(col_filter[factor(chr_raw$filter)], 90),
     peakType = "none", main = "BPC")
```

# Peak detection

```{r peak-detection}
cwp <- CentWaveParam(ppm = 20,
                     peakwidth = c(2, 20),
                     prefilter = c(5, int),
                     snthresh = 5,
                     noise = 1000,
                     mzdiff = 0.001,
                     integrate = 2)
xdata <- findChromPeaks(data_raw, param = cwp)
```


## Filter low-intensity peaks

```{r filter-intensity}
xdata <- refineChromPeaks(xdata, 
                          param = FilterIntensityParam(
                            nValues = 5, threshold = int)
)
```


## Peak post-processing

```{r peak-postproc}
mnp <- MergeNeighboringPeaksParam(
  expandRt = 2, 
  expandMz = 0.001, 
  ppm = 10,
  minProp = 0.66)
register(SerialParam())
xdata <- refineChromPeaks(xdata, param = mnp)
```


# Aligment

```{r aligment, eval = FALSE}
# Define the hook peaks
pdp <- PeakDensityParam(sampleGroups = xdata$type,
                        minFraction = 0.75,
                        binSize = 0.02,
                        bw = 3)
xdata <- groupChromPeaks(xdata, param = pdp)
# Perform the aligment
pgp <- PeakGroupsParam(span = 0.3,
                       subset = which(xdata$type == "QC"),
                       subsetAdjust = "average",
                       minFraction = 1)
xdata <- adjustRtime(xdata, param = pgp)
```


## Checking

The results from the alignment are shown below. 
To visualize the alignment results, the BPC for the whole data set is plotted.

```{r alignment-rtime-plot, fig.width = 12, fig.height = 10, eval = FALSE}
chr_raw <- chromatogram(xdata, aggregationFun = "max", 
                        adjustedRtime = FALSE)
chr_adj <- chromatogram(xdata, aggregationFun = "max")
par(mfrow = c(3, 1), mar = c(0, 4.3, 2, 0.1))
plot(chr_raw, col = paste0(col_filter[factor(chr_raw$type)], 40),
     peakType = "none", main = "BPC, raw")
plot(chr_adj, col = paste0(col_filter[factor(chr_raw$type)], 40),
     peakType = "none", main = "BPC, adjusted")
plotAdjustedRtime(xdata, col = paste0(col_filter[factor(xdata$type)],80))
```


# Correspondance

```{r correspondance}
pdp <- PeakDensityParam(sampleGroups = xdata$solvent,
                        minFraction = 0.25,
                        minSamples = 2,
                        binSize = 0.02,
                        bw = 1)
xdata <- groupChromPeaks(xdata, param = pdp)
```


# Peak filling

```{r peak-filling}
fcp <- ChromPeakAreaParam()
xdata <- fillChromPeaks(xdata, param = fcp)
```


```{r peak-filling-checking-correlation, eval = FALSE}
data <- quantify(xdata, method = "sum", filled = FALSE)
assays(data)$raw_filled <- featureValues(xdata, method = "sum",
                                         filled = TRUE)
## Select only the filled-in signals
tmp <- assay(data, "raw_filled")
tmp[!is.na(assay(data, "raw"))] <- NA
assays(data)$raw_only_filled <- tmp
avg_det <- rowMeans(assay(data, "raw")[, data$type == "QC"], na.rm = TRUE)
avg_fil <- rowMeans(assay(data, "raw_only_filled")[, data$type == "QC"],
                    na.rm = TRUE)
plot(log2(avg_det), log2(avg_fil), xlab = "detected", ylab = "filled-in",
     main = "Feature abundances, QC samples", pch = 16, col = "#00000080")
abline(0, 1, col = "grey")
cor(log2(avg_det), log2(avg_fil), use = "pairwise.complete.obs")
```

We can see a relatively high correlation between the detected and filled-in
signal.

Below, we are going to plot the outliers observed in the previous plot (in case there is any outlier).

```{r peak-filling-ckecking-eic, eval = FALSE}
fts <- names(which(log2(avg_det) > (log2(avg_fil)*1.5)))
if(length(fts) > 0){
  for(i in 1:length(fts)){
    chrs <- featureChromatograms(xdata, 
                                 features = fts[i], 
                                 expandRt = 10, filled = FALSE)
    pk_col <- col_filter[as.character(
      chrs$class[chromPeaks(chrs)[, "sample"]])]
    
    plotChromPeakDensity(chrs, peakPch = 16, 
                         peakCol = paste0(pk_col, 80),
                         peakBg = paste0(pk_col, 10),
                         simulate = FALSE)
    
    chrs <- featureChromatograms(xdata, 
                                 features = fts[i], 
                                 expandRt = 10, filled = TRUE)
    pk_col <- col_filter[as.character(
      chrs$class[chromPeaks(chrs)[, "sample"]])]
    
    plotChromPeakDensity(chrs, peakPch = 16, 
                         peakCol = paste0(pk_col, 80),
                         peakBg = paste0(pk_col, 10),
                         simulate = FALSE)
  }
}
```

# PCA

```{r}
data <- featureValues(xdata, method = "sum", value = "into")
set.seed(123)
data <- t(imputeRowMinRand(data, method = "from_to",
                           min_fraction = 1/100,
                           min_fraction_from = 1/1000
))
dt <- log10(data)
scaling.pareto <- scalefun(sc.p="pareto")
dt <- data.frame(scaling.pareto(dt))
pca <- prcomp(dt, center = FALSE, scale. = FALSE)

tmp <- data.frame(pca$x)
plot_ly(x = tmp$PC1, y = tmp$PC2,
        text = gsub(".mzData", "", rownames(tmp)),
        color = xdata$solvent,
        symbol = xdata$filter)
plot_ly(x = tmp$PC1, y = tmp$PC2,
        text = gsub(".mzData", "", rownames(tmp)),
        color = xdata$solvent)
plot_ly(x = tmp$PC1, y = tmp$PC2,
        text = gsub(".mzData", "", rownames(tmp)),
        color = xdata$filter)

features <- data.frame(featureDefinitions(xdata))
tmp <- data.frame(pca$rotation)
rownames(tmp) <- paste(sprintf("%.4f",round(features$mzmed, 4)), 
                       sprintf("%.2f",round(features$rtmed/60, 2)), 
                       sep = "_")
plot_ly(x = tmp$PC1, y = tmp$PC2,
        text = rownames(tmp))
```

# Save data

```{r save}
save(xdata, file = paste0("data/RData/data_XCMS_", polarity, ".RData"))
```


# Session information

```{r session}
Sys.time()-startpoint
devtools::session_info()
```